Transitivity of Subtyping for Intersection Types

### 2020-05-14 papers/LMCS-subtyping-intersections/paper.tex
finished draft of new motivation section

\begin{abstract}
  The subtyping rules for intersection type systems traditionally
  employ a transitivity rule (Barendregt et al. 1983), which means
  that subtyping does not enjoy the subformula property.  Laurent
  develops a sequent-style subtyping system, without transitivity, and
  proves transitivity via a sequence of six lemmas that culminate in
  cut-elimination (2018). This article develops a subtyping system in
  regular style that omits transitivity and provides a direct proof of
  transitivity, significantly reducing the length of the proof. The
  new system satisfies something like the subformula property: every
  type occuring in the derivation of $A <: B$ is a subformula of $A$
  or $B$, or an intersection of such subformulas.  Borrowing from
  Laurent's system, the rule for function types is essentially the
  $\beta$-soundness property.  The main lemma required for the
  transitivity proof is one that has been used to prove the inversion
  principle for subtyping of function types. The article concludes
  with a proof that the new subtyping system is equivalent to that of
  Barendregt, Coppo, and Dezani-Ciancaglini.
\end{abstract}

@@

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle as a tool for studying normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
been used for at least three purposes: to enhance the expressiveness
of type systems
~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
to increase the precision of static
analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
and to express filter models for the denotational semantics of a
variety of lambda calculi
~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.

The work in this

A filter model

\[
   \llbracket M \rrbracket = \{ A \mid \emptyset \vdash M : A \}
\]

Perhaps the best-known  them is the BCD intersection type system of
\citet{Barendregt:1983aa}. For this article we focus on the BCD
system, following the presentation of \citet{Barendregt:2013aa}.  We
conjecture that our results apply to other intersection type systems
that include the $({\to}{\cap})$ rule for distributing intersection
and function types.

The BCD intersection type systems extends the simply-typed lambda
calculus with the addition of intersection types, written $A \cap B$, a
top type $\TOP$, and an infinite collection of type
constants. Figure~\ref{fig:types} defines the grammar of types.

@@

The BCD intersection type system includes a subsumption rule which
states that a term $M$ in environment $\Gamma$ can be given type $B$
if it has type $A$ and $A$ is a subtype of $B$, written $A \leq B$.
\[
\inference{\Gamma \vdash M : A & A \leq B}
          {\Gamma \vdash M : B}
\]
Figure~\ref{fig:BCD-subtyping} reviews the BCD rules for subtyping.
Note that in the (trans) rule, the type $B$ appears in the premises
but not in the conclusion. Thus, the BCD subtyping judgment does not
enjoy the subformula property.  For many other subtyping systems, it
is straightforward to remove the (trans) rule, modify the other rules,
and then prove transitivity.  Unfortunately, the $({\to}{\cap})$
rule of the BCD system significantly complicates the situation.

@@

The subformula property is a useful one. For example, the author uses
intersection types to create denotational semantics (a filter model)
for the ISWIM language, which includes constants and primitive
operations~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
This requires placing extra conditions on types and it is much easier
to do so when subtyping satisfies the subformula property.

There has been a recent flurry of interest in subtyping for
intersection types that has led to several systems that satisfy the
subformula property. We discuss each of them in the next few
paragraphs.

\citet{Laurent:2018aa} introduces the ISC sequent-style system,
written $\Gamma \vdash B$, where $\Gamma$ is a sequence of types
$A_1,\ldots,A_n$. The intuition is that $A_1,\ldots,A_n \vdash B$
corresponds to $A_1 \cap \cdots \cap A_n \leq B$. The ISC system satisfies the
subformula property and is equivalent to the BCD system. To prove
this, Laurent establishes six lemmas that culminate in
cut-elimination, from which transitivity follows.

\citet{Bi:2018aa} present a subtyping algorithm that satisfies the
subformula property and prove that is equivalent to BCD
subtyping. Their system is based on the decision procedure
of~\citet{Pierce:1989aa} and takes the form $\mathcal{L} \vdash A
\prec: B$, where $\mathcal{L}$ is a queue of types that are peeled off
from the domain of the type on the right. So the judgement
$\mathcal{L} \vdash A \prec: B$ is equivalent to $A \leq \mathcal{L}
\to B$ where $\mathcal{L} \to B$ is defined as
\begin{align*}
  [] \to B &= B\\
  (\mathcal{L},A) \to B &= \mathcal{L} \to (A \to B)
\end{align*}
The proof of transitivity for this system is an adaptation of Pierce's
but it corrects some errors and introduces more lemmas concerning an
auxilliary notion called reflexive supertypes.

\citet{Muehlboeck:2018aa} develop a framework for obtaining subtyping
algorithms for systems that include intersection and union types and
that are extensible to other types. To decide $A \leq B$ their
algorithm converts $A$ to disjunctive normal form and then applies a
client-supplied function to each collection of literals. To obtain a
system equivalent to BCD, including the ${\to}{\cap}$ rule, the client
side function saturates the collection of literals by applying
${\to}{\cap}$ (left to right) as much as possible.
[todo: say something about the proof of transitivity]

This article presents a direct route to the subformula property and
proof of transitivity for BCD subtyping. We present straightforward
rules for a new subtyping relation $A <: B$ and directly prove
transitivity. The key to $A <: B$ is a rule for function types similar
to the one in Laurent's ISC, which effectively turns the
$\beta$-soundness property~\citep{Barendregt:2013aa} into a subtyping
rule. While the rules that define $A <: B$ are not particularly novel,
the fact that transivity can be proved directly from them is
suprising.

We describe the new subtyping relation in
Section~\ref{sec:new-subtyping}, prove transitivity in
Section~\ref{sec:trans}, and prove its equivalence to BCD subtyping in
Section~\ref{sec:equiv}. We make some concluding remarks in
Section~\ref{sec:conclude}.

The definitions and results in this article have been machine checked
in Agda.

>>>>

\begin{abstract}
  The subtyping rules for intersection type systems traditionally
  employ a transitivity rule (Barendregt et al. 1983), which means
  that subtyping does not satisfy the subformula property, making it
  more difficult to use in filter models for compiler verification.
  Laurent develops a sequent-style subtyping system, without
  transitivity, and proves transitivity via a sequence of six lemmas
  that culminate in cut-elimination (2018). This article develops a
  subtyping system in regular style that omits transitivity and
  provides a direct proof of transitivity, significantly reducing the
  length of the proof. The new system satisfies something like the
  subformula property: every type occuring in the derivation of $A <:
  B$ is a subformula of $A$ or $B$, or an intersection of such
  subformulas.  Borrowing from Laurent's system, the rule for function
  types is essentially the $\beta$-soundness property.  The main lemma
  required for the transitivity proof is one that has been used to
  prove the inversion principle for subtyping of function types. The
  article concludes with a proof that the new subtyping system is
  equivalent to that of Barendregt, Coppo, and Dezani-Ciancaglini.
\end{abstract}

@@

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle to study normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection type systems
have been used for at least three purposes:
\begin{itemize}
  \item as type systems for programming
    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
  \item to formulate precise static
    analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
    and
  \item to express filter models for the denotational semantics of a
    variety of lambda
    calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
\end{itemize}

There are many intersection type systems; perhaps the best-known them
is the BCD intersection type system of \citet{Barendregt:1983aa}. For
this article we focus on the BCD system, following the presentation of
\citet{Barendregt:2013aa}. We conjecture that our results apply to
other intersection type systems that include the $({\to}{\cap})$ rule
for distributing intersection and function types.

The BCD intersection type system includes function types, $A \to B$,
intersection types, $A \cap B$, a top type, $\TOP$, and an infinite
collection of type constants. Figure~\ref{fig:types} defines the
grammar of types.

@@

The BCD intersection type system includes a subsumption rule which
states that a term $M$ in environment $\Gamma$ can be given type $B$
if it has type $A$ and $A$ is a subtype of $B$, written $A \leq B$.
\[
\inference{\Gamma \vdash M : A & A \leq B}
          {\Gamma \vdash M : B}
\]
Figure~\ref{fig:BCD-subtyping} reviews the BCD subtyping system.  Note
that in the (trans) rule, the type $B$ appears in the premises but not
in the conclusion. Thus, the BCD subtyping system does not satisfy the
subformula property.  For many other subtyping systems, it is
straightforward to remove the (trans) rule, modify the other rules,
and then prove transitivity.  Unfortunately, the $({\to}{\cap})$ rule
of the BCD system significantly complicates the situation.

@@

\section{Motivations from Verified Compilation}

Our interest in intersection type systems stems from our work on
mechanized proofs of correctness for compiling functional programming
languages. We are exploring whether the use of denotational semantics
can streamline such proofs compared to the operational techniques,
e.g., used to prove the correctness of
CakeML~\citep{Kumar:2014aa,Owens:2017aa}. In particular, we are
constructing filter models (based on intersection type systems)
because they are relatively straightforward to mechanize in a proof
assistant compared to traditional domain
theory~\citep{Benton:2009ab,Dockins:2014aa}.  In a filter model, the
meaning of a program $M$ is the set of all types that can be assigned
to it by an intersection type system:
\[
   \llbracket M \rrbracket = \{ A \mid \emptyset \vdash M : A \}
\]
Most type systems are much too coarse of an abstraction to be used as
filter models. Intersection type systems are special in that they can
completely characterize the runtime behavior of a program.  To catch a
glimpse of how filter models work, consider the meaning of the
identity function in the $\lambda$-calculus.
\[
  \llbracket \lambda x.\, x \rrbracket =
      \{ c_0 \to c_0, c_1 \to c_1, c_2 \to c_2, \ldots \}
\]
If one takes each type constant $c_i$ to be a singleton type for the
number $i$, then the set of types assigned to the identity function
starts to look like the graph of the function. Indeed, filter models
are closely related to graph
models~\citep{Scott:1976lq,Engeler:1981aa,Plotkin:1993ab}.

One of the standard compiler transformations that we are interested in
verifying is common subexpression elimination. For example, this
transformation would replace the duplicated $(f \; x)$ terms in the
following
\[
  M = \lambda f.\, \lambda h.\, \lambda x.\, h \; (f \; x) \; (f \; x)
\]
with a single instance of $(f\;x)$ that is let-bound to a variable $y$:
\[
M' = \lambda f.\, \lambda h.\, \lambda x.\,
\mathsf{let}\; y = (f \; x) \;\mathsf{in}\;
h \; y \; y
\]
Unfortunately, $M$ and $M'$ are not denotationally equal according to
most filter models because filter models allow function graphs that
represent arbitrary relations. The above transformation depends on $(f
\; x)$ producing the same result every time, but a filter model
associates arbitrary graphs with parameter $f$.

This problem is not conceptually difficult to solve; one just needs to
restrict the graphs to only allow approximations of functions. This
can be accomplished by requiring every entry in a function's graph to
be ``consistent''. Two function types, $A \to B$ and $A' \to B'$, are
consistent, written $(A \to B) \sim (A' \to B')$, when either $A
\not\sim A'$ or $A \sim A'$ and $B \sim B'$. Two type constants are
consistent when they are equal. One would then hope to prove that for
any term $M$, if $\Gamma \vdash M : C$ and $\Gamma \vdash M : C'$,
then $C \sim C'$ as long as $C$, $C'$, and $\Gamma$ are self
consistent. A key lemma needed for that proof is that consistency is
upward closed with respect to subtyping: if $A \sim B$, $A <: C$, $B
<: D$, then $C \sim D$.

This brings us back to the transitivity rule and the subformula
property. The proof of this lemma hits a snag in the cases where $A <:
C$ and $B <: D$ are derived using the transitivity rule. The
intermediate type is not guaranteed to be self consistent. We explored
adding this guarantee to the transitivity rule, but it caused a
significant increase in obligations elsewhere in the proof. By some
serendipity, around that moment in time \citet{Laurent:2018aa}
published the first subtyping system for intersection types that
removes the transitivity rule and satisfies the subformula property.
Laurent developed a sequent-style subtyping system, written $\Gamma
\vdash B$, where $\Gamma$ is a sequence of types $A_1,\ldots,A_n$. The
intuition is that $A_1,\ldots,A_n \vdash B$ corresponds to $A_1 \cap
\cdots \cap A_n \leq B$. This system satisfies the subformula property
and is equivalent to the BCD system. To prove this, Laurent
establishes six lemmas that culminate in cut-elimination, from which
transitivity follows. Laurent mechanized these proofs in Coq.

We immediately took advantage of Laurent's result to complete a filter
model for the ISWIM
language~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
We mechanized the formal development and proofs in
Isabelle~\citep{Siek:2018aa}.

A year later we began to port this filter model to the Agda proof
assistant, motivated by the need to use dependent types to make our
development more reusable: we needed several variations on the filter
model to give semantics to different languages, including ones that
are call-by-value and call-by-name, and to the different intermediate
languages used in our compiler. We found it tedious to repeat
Laurent's proof of transitivity for each of these variations, and in
the process developed an intuition that the sequent-style system was
not necessary. The present article discusses how we were able to
extract the key feature from Laurent's system, the rule for subtyping
between function types (which effectively turns the $\beta$-soundness
property~\citep{Barendregt:2013aa} into a subtyping rule) and splice
it into the BCD subtyping system, replacing the function $(\to)$,
distributivity $({\to}{\cap})$, and transitivity (trans) rules.  We
then prove transitivity directly, without the sequence of six lemmas,
instead using one lemma from the literature on inversion principles
for subtyping of function types~\citep{Barendregt:2013aa}. The end
product is a subtyping system that highly suitable for mechanizing
filter models.

We discuss two recent developments regarding subtyping for
intersection types in Section~\ref{sec:other-developments}.  We
present our new subtyping system in Section~\ref{sec:new-subtyping},
prove transitivity in Section~\ref{sec:trans}, and prove its
equivalence to BCD subtyping in Section~\ref{sec:equiv}. We make some
concluding remarks in Section~\ref{sec:conclude}.

The definitions and results in this article have been machine checked
in Agda, in the file \texttt{agda/TransSubInter.agda} in the following
repository.

\begin{center}
  \url{https://github.com/jsiek/denotational_semantics}
\end{center}

\section{Other Developments}
\label{sec:other-developments}

There has been a recent flurry of interest in subtyping for
intersection types that has led to two more systems that satisfy the
subformula property, in addition to the one by \citet{Laurent:2018aa}
discussed above. These two systems were developed for the purposes of
type checking programming languages, and not for the construction of
filter models, so the goals are somewhat different.

\citet{Bi:2018aa} present a subtyping algorithm that satisfies the
subformula property and proves that it is equivalent to BCD
subtyping. Their system is based on the decision procedure
of~\citet{Pierce:1989aa} and takes the form $\mathcal{L} \vdash A
\prec: B$, where $\mathcal{L}$ is a queue of types that are peeled off
from the domain of the type on the right. So the judgement
$\mathcal{L} \vdash A \prec: B$ is equivalent to $A \leq \mathcal{L}
\to B$ where $\mathcal{L} \to B$ is defined as
\begin{align*}
  [] \to B &= B\\
  (\mathcal{L},A) \to B &= \mathcal{L} \to (A \to B)
\end{align*}
The proof of transitivity for this system is an adaptation of Pierce's
but it corrects some errors and introduces more lemmas concerning an
auxilliary notion called reflexive supertypes.

\citet{Muehlboeck:2018aa} develop a framework for obtaining subtyping
algorithms for systems that include intersection and union types and
that are extensible to other types. To decide $A \leq B$ their
algorithm converts $A$ to disjunctive normal form and then applies a
client-supplied function to each collection of literals. To obtain a
system equivalent to BCD, including the ${\to}{\cap}$ rule, the client
side function saturates the collection of literals by applying
${\to}{\cap}$ (left to right) as much as possible.
%[todo: say something about the proof of transitivity]
